---
title: "Week 2 Gated Flow Data"
author: "Christine Wiggins"
date: "4/24/2023"
output: html_document
---

```{r setup, include=FALSE}
#rm(list = ls())
knitr::opts_chunk$set(echo = TRUE)
```
Load packages
```{r load}
library(systemsseRology)
#library(readxl)
library(pheatmap)
library(viridis)
library(RColorBrewer)
library(summarytools)
library(ggpubr)
#set.seed(1234)  # random seed for reproducibility
# Create folder if it does not exist
if (!(file.exists("Figures"))) {
dir.create(file.path(getwd(), "Figures"))
}
```
# Classification and preprocessing

Read the data from an Excel file. Z-score the data and assign labels in the variables y1 and y2.

Data include flow cytometry measurements for 52 mice across 2 phenotypes (NCan, Canonical)


```{r data}
#load and format data
data <- as.data.frame(read.csv("Data/consolidated_2wk_master.csv"))
data <- data[,-1]
ID_wk2 <- data$uniqueID
X <- suppressWarnings(as.matrix(as.data.frame(lapply(data[1:nrow(data), 8:ncol(data)], as.numeric))))
X <- as.data.frame(X)
rownames(X) <- data$uniqueID
colnames(X) <- colnames(data)[8:ncol(data)]
X$ID <- ID_wk2


#Create set of outcomes to later regress on (lung and spleen CFU count)
#gender <- factor(data[1:nrow(data),3])

outcomenames <- c("Spleen_CFU","Lung_CFU_Log10","Percent_weight_change",
                  "Weight_Start","Weight_2wk",
                  "Total_IFNg_p300")
outcomes <- X[, (colnames(X) %in% outcomenames)]

#Load updated CFU, cytokine values from replated experiment
data_cfu2 <- as.data.frame(read_excel("Data/2 week/Chal74 2wk data additions.xlsx"), header = T)

lung_cyt2_raw <- as.data.frame(read_excel("Data/2 week/Chal74 2wk data additions.xlsx"), 
                               header = T)[,1:19]

#Take only outcomes of interest
data_cfu2 <- data_cfu2[,c(1,3,4,20)]
#Create vector of phenotype, 
data_cfu2$Group[which(grepl("RSTR",data_cfu2$Group))] <- "R"
data_cfu2$Group[which(grepl("Canonical",data_cfu2$Group))] <- "C"

#Generate unique ID for each mouse
data_cfu2$uniqueID <- paste0(data_cfu2$Genotype,"_",
                               data_cfu2$Group, "_",
                               data_cfu2$'Mouse ID')

#Add outcomes back to flow cytometry data
X <- merge(x=X,y=data_cfu2, by.x = "ID", by.y = "uniqueID")
X <- X[, !(colnames(X) %in% outcomenames)]
X[is.na(X)] <- 0

#Create vector of mouse genotypes
y1_wk2 <- X$Genotype
y1_wk2 <- sub("_.*", "", y1_wk2)
y1_wk2 <-factor(y1_wk2)
y1_wk2 <- droplevels(y1_wk2)
y1_wk2 <- as.factor(gsub("J","",y1_wk2))

#Create vectors of phenotypes, CFU, and ID
y2_wk2 <- factor(X$Group)
cfu_wk2 <- X$Lung_CFU_Log10.y
ID_wk2 <- X$ID

#Remove metadata from flow cytometry data
X <- X[,-which(grepl("Group|Lung_CFU_Log10|ID|Genotype|Weight|Total_Cells", colnames(X)))]
df_wk2 <- as.data.frame(X)

#remove the fully no cytokine information subset
df_wk2 <- df_wk2[,-which(colSums(df_wk2) < .00001)]

```

#Z-scoring

```{r Z-score}
#Zscore data
X <- as.data.frame(scale(df_wk2, center = TRUE, scale = TRUE))
cfu <- as.matrix(scale(cfu_wk2, center = TRUE, scale = TRUE))
rownames(X) <- rownames(df_wk2)

#Repeat only for canonical
X_canon <- as.data.frame(scale(df_wk2[which(y2_wk2 == "C"),], center = TRUE, scale = TRUE))
cfu_canon <- as.matrix(scale(cfu_wk2[which(y2_wk2 == "C")], center = TRUE, scale = TRUE))
rownames(X_canon) <- rownames(df_wk2[which(y2_wk2 == "C"),])

#Repeat only for NCAN
X_NCAN <- as.data.frame(scale(df_wk2[which(y2_wk2 == "R"),], center = TRUE, scale = TRUE))
cfu_NCAN <- as.matrix(scale(cfu_wk2[which(y2_wk2 == "R")], center = TRUE, scale = TRUE))
rownames(X_NCAN) <- rownames(df_wk2[which(y2_wk2 == "R"),])
```

The variable df_features contains information about the features, mainly used for visualization routines. 

```{r annotation}
library(tidyr)
library(stringr)
#Pull out identifying number from Excel sheet subject name
identity <- rownames(X)

#Create data frame with corresponding identity, strain, and resistor status
df_id <- data.frame(identity = identity, strain = factor(y1_wk2), 
                    status = factor(y2_wk2))
print(df_id)

df_features <- data.frame(name = colnames(X))
df_features$label <- factor(df_features$name)
print(df_features)
```

#PLSDA
#Feature Selection Validation and Permutation testing

```{r Select Features}
library(ropls)
library(systemsseRology)

# general options for plotting
my_colors <- list(
  status = c("R" = '#8C02CC', "C" = '#5192C3', "equal" = 'black'), 
  strain = c(viridis(9))
  )
names(my_colors$strain) <- c("CC009", "CC024","CC039", "PWK", "CC011", "CC059", "CC001",  "B6", "CC018")

#select features with LASSO
#Note: because LASSO involves partitioning the data, you can get slightly different
#feature results on each run.
opts_sel <- list(n_trials = 100, threshold = 0.8, return_count = TRUE)
sel_features <- select_repeat(as.matrix(X), y2_wk2, 
                              selector = select_lasso, 
                              options = opts_sel)
X_sel <- X[, sel_features$sel_features]

# Perform a PLS-DA using the selected features and plot the scores and loadings
# Check number of latent variables and increase to 2 if <2 (for visualization purposes)
opts_plot <- list(df_id = df_id,
                  loading_alpha = 1, # transparency for the loadings
                  score_alpha = 1, # transparency for the scores
                  LV_ind = c(1,2), # which LVs to plot
                  color_features = "status", # according to which property the features should be color-coded
                  colors = my_colors,
                  y_name = "status") 

opts_plot2 <- list(df_features = df_features,
                  loading_alpha = 1, # transparency for the loadings
                  score_alpha = 1, # transparency for the scores
                  LV_ind = c(1,2), # which LVs to plot
                  colors = my_colors,
                  y_name = "status")

opts_model <- list(n_LV = 2)

#Use only the LASSO-selected features to build the model
model <- train_ropls(X_sel, y2_wk2, options = opts_model)
#print model summary- check R2X and R2Y to see that they're reasonable
ropls::getSummaryDF(model)

#Visualize the scores and loadings
plt_scores <- visualize_ropls_scores(model, y2_wk2, options = opts_plot)
print(plt_scores)

plt_loadings <- visualize_ropls_loadings(model, options = opts_plot2)
print(plt_loadings)

# set additional options required to color code enrichment in the bar plot of the loadings
opts_plot2$X <- X_sel
opts_plot2$y <- y2_wk2
opts_plot2$LV_ind <- 1
opts_plot2$mark_enrichment <- TRUE
plt_loadings_bar1 <- visualize_ropls_loadings_bar(model, options = opts_plot2)
print(plt_loadings_bar1)
opts_plot2$LV_ind <- 2
plt_loadings_bar2 <- visualize_ropls_loadings_bar(model, options = opts_plot2)
print(plt_loadings_bar2)

```

#Validation, Permutation Testing on Selected Features

```{r Validation, Permutation of Selected Features}
#How many rounds of feature selection do you want to run? What % of 
#time should a feature be selected in order to be used (here is 80%)?
opts_sel <- list(n_trials = 100, threshold = 0.8, return_count = FALSE)
select <- function(X, y) { return(select_repeat(X, y, selector = select_lasso, options = opts_sel)) }

#Input feature selection method, model training method, prediction method,
#and how you want your results scores to be calculated
method = list(select = select,
              train = train_ropls,
              predict = predict_ropls,
              score = score_accuracy)

#How many folds of cross-validation do you want, and how many random feature
#and permuted lable trials?
opts = list(n_folds = 5, pt_trials = 10, rf_trials = 10)
return_vals2 <- cross_validation_unpaired(as.matrix(X), 
                                          y2_wk2, method, opts, n_trials = 10)
plt <- visualize_validate(return_vals2)
u2 <- rowMeans(sapply(return_vals2, unlist))[1]
u2
u3 <- rowMeans(sapply(return_vals2,unlist))
u3

# pdf("../Sassetti TB collaboration/crossVal_2wk_canon_v_NCAN_new_gating.pdf", width = 8, height = 5) 
# print(plt)
# dev.off()

#Pull 5-fold CV, random feature, and permuted labels scores together
#you'll need to calculate p-value yourself & not rely on the automatic one in the plot
validation_df <- data.frame(cv = unlist(return_vals2)[which(names(unlist(return_vals2)) == "cv_score")])
null_df <- data.frame(rf = unlist(return_vals2)[which(grepl("rf_scores", names(unlist(return_vals2))))],
                  pt = unlist(return_vals2)[which(grepl("pt_scores", names(unlist(return_vals2))))])

#Save data
# write.csv(validation_df, file = "../crossvalidation_2_wk_canon_v_NCAN_new_gating.csv")
# write.csv(null_df, file = "../null_2_wk_canon_v_NCAN_new_gating.csv")

```


#Correlation Mapping

```{r correlation heatmap}
library(RColorBrewer)
corrMat <- cor(X, method = "spearman")
# heatmap for full correlation matrix
pheatmap(-corrMat, cluster_rows = FALSE, cluster_cols = FALSE,        
         color = colorRampPalette(brewer.pal(n = 10, name = "RdBu"))(100),
         breaks = seq(1, -1, length = 100),
         cellwidth = .5, cellheight = .5, fontsize = 6, border_color = "NA") 

# keep only selected features and highly correlated features
colMax <- function(X) apply(X, 2, max)
corrMat <- corrMat[sel_features$sel_features, ]
corrMat <- corrMat[, which(colMax(abs(corrMat)) < 1 & colMax(abs(corrMat)) > 0.7)]
#recorder corrMat to group NCAN and canonical features together
corrMat <- corrMat[c("Per_CD4_FOXP3p_Tbetn","CD4_CD39n_CD25p","Per_DC_CD11b",
                     "CD4_CCR6p_CXCR3n","MCP1","Per_PMN_Nos2n_IL1bp",
                     "CD4_CD44p_CD26Ln","IL7","Per_cd8_IL4_p300","Per_cd8_IL4_CFPMtb",
                     "Per_DC_CD11b_PDL1n_CD14p",
                     "nCD40LnCD107anIFNgnIL2nIL4nIL10nIL17.TNF",
                     "CD4_FOXP3n_Tbetp_CXCR3p_RORgtn","DN_GATA3n_RORgtp",
                     "CD40LnCD107anIFNgnIL2nIL4nIL10nIL17.TNF",
                     "Per_Macrophage_PDL1n_CD14p"), ]

# heatmap for reduced correlation matrix
plt <-pheatmap(-corrMat, cluster_rows = FALSE, cluster_cols = TRUE,    
         color = colorRampPalette(brewer.pal(n = 10, name = "RdBu"))(100),
         breaks = seq(1, -1, length = 100),
         treeheight_col = 0, treeheight_row = 0,
         cellwidth = 6, cellheight = 8, fontsize = 8, border_color = "NA")
# pdf("Figures/New gating figures/cocorrelate_network_2wk_new gating.pdf", width = 20, height = 8) 
# print(plt)
# dev.off()
```
