---
title: "COMPASS_clean"
author: "Christine Wiggins"
date: "2024-04-29"
output: html_document
---

#to run this, you need the csv files exported
#from FlowJo that have polyfunctional subset frequencies, as well as parent
#you can generate these from FCS files directly

#subset cell counts and metadata
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(reshape2)
library(COMPASS)
library(ggplot2)
library(dplyr)
library(reshape2)
#library(readxl)
library(readr)
library(stringr)
library(gridExtra)
library(tidyverse)
library(ggpubr)
```

#To start from FlowJo csv files, you need:
1) frequency of polyfunctional subsets (exclude bulk & single cytokine)
2) parent cell count total
3) count data for polyfunctional subsets (can get via multiplying 1 & 2 and rounding)

You need this for both your stimulated and unstimulated conditions

```{r load from FlowJo}
#load count data for the stimulated and unstimulated (combined in one)
compass_data <- as.data.frame(read.csv('COMPASS/CD4_p300_data.csv',
                                       header=TRUE, check.names = FALSE,
                                       stringsAsFactors = FALSE))
rownames(compass_data) <- compass_data[,1]
compass_data <- compass_data[,2:ncol(compass_data)]

#################### METADATA #########################

metadata <- read.csv('COMPASS/CD4_p300_metadata.csv', 
                     header=TRUE, check.names = FALSE)
metadata <- as.data.frame(metadata)
rownames(metadata) <- metadata[,1]
metadata <- metadata[,2:ncol(metadata)]

################# remove any missing data #################

metadata <- metadata[-which(metadata$unique_id == names(which(is.na(rowMeans(compass_data))))),]
compass_data<- compass_data[-which(grepl(names(which(is.na(rowMeans(compass_data)))), rownames(compass_data))),]

######### Split data into stimulated and unstimulated ##########

n_u = as.matrix(subset(compass_data, metadata$stim == "unstim"))
n_s = as.matrix(subset(compass_data, metadata$stim == "p300"))

####Name and order rows of all the matrices #############

# assign consistent row names to n_u and n_s
rownames(n_u) = subset(metadata$unique_id, metadata$stim=="unstim")
rownames(n_s) = subset(metadata$unique_id, metadata$stim=="p300")
# Now all matrices have the same dimensions and appropriate rownames

#run compass
compass_fit = COMPASS::SimpleCOMPASS(n_s = n_s, 
                                      n_u = n_u,
                                      meta = subset(metadata, 
                                                    metadata$stim == "unstim"), 
                                     #just take one metadata, same for stim/unstim
                                      individual_id = "unique_id", 
                                      iterations = 40000, #n_iterations
                                      replications = 8)

```

#To start from a saved COMPASS object
```{r load from COMPASS}
#load correct COMPASS object

compass_fit <- readRDS("your_compass_object.RDS")
#to run this for the paper, use provided cd4_p300_wk5_fit.RDS

```

#Change below to use whatever metadata is relevant to your plotting
#For paper, we had phenotype info in the COMPASS metadata
```{r polyfunctionality plots}
#determining significance for pfs plot
pfs <- as.data.frame(PolyfunctionalityScore(compass_fit))
pfs$Phenotype <- compass_fit$data$meta$Phenotype
re <- pfs[which(pfs$Phenotype == "R"), 1]
ca <- pfs[which(pfs$Phenotype == "C"), 1]
p <- wilcox.test(re, ca)[["p.value"]]

#plot polyfunctionality scores
pfs_plot <- ggplot(data = scores(compass_fit), 
                   mapping = aes(x = Phenotype, y = PFS, fill = Phenotype)) +
    geom_boxplot(outlier.colour = NA) +
    geom_jitter() +
    stat_summary(fun = median, geom = "text", aes(label = after_stat(y)), vjust = 15) + 
    theme_classic() +
    scale_fill_manual(breaks = c("R", "C"), values = c('#8C02CC', '#5192C3')) +
    labs(y = "Polyfunctionality Score") + 
    theme(plot.caption = element_text(vjust = 160, hjust = 0.7)) + 
    labs(caption = paste0("p = ", p)) 

#save pdf
pdf("your_pdf.pdf", width = 5, height = 7)
print(pfs_plot)
dev.off()

#save data
library(openxlsx)
write.xlsx(pfs, "Data_for_stim.xlsx")
```

#Get heatmaps of all subsets detected to be antigen-specific
```{r all detected subset heatmaps}

#use compass plotting
plot(compass_fit, "Phenotype")

pdf("your_pdf.pdf", width = 20, height = 20, paper = "letter")
plot(compass_fit, "Phenotype")
dev.off()
```

#Downselect to only subsets different between NCan & canonical mice & above
#a certain threshold
```{r sig diff subsets}
#saving the filtered subsets that COMPASS identifies
threshold = 0.01
means <- apply(compass_fit$fit$mean_gamma, 2, function(x) {
  mean(x)
})
i <- means > threshold
filtered_subsets <- names(which(i))

#Pull data from COMPASS object
corrected_dat <- compass_fit$data$n_s #stimulated data
meta <- compass_fit$data$meta

```

#Plot % frequencies - need all the original polyfunctional frequencies from
#FlowJo exports for stim & unstim conditions
```{r subset frequency plotting}

#want frequencies
cd4_p300_data <- as.data.frame(read_csv("Count Frequency CD4 p300.csv"))
cd4_unstim_data <- as.data.frame(read_csv("Count Frequency CD4 unstim.csv"))
cd4_p300_parent_count <- as.data.frame(read_csv("Parent Count CD4 p300.csv"))
cd4_unstim_parent_count <- as.data.frame(read_csv("Parent Count CD4 unstim.csv"))

#take out mean, sd from csvs from FlowJo
#This is hard-coded, so double check row & column indices
cd4_p300_data <- as.data.frame(cd4_p300_data[1:55,2:(ncol(cd4_p300_data)-1)])
cd4_unstim_data <- as.data.frame(cd4_unstim_data[1:55,2:(ncol(cd4_unstim_data)-1)])

#also need metadata so i can take out FMOs & controls
metadata <- as.data.frame(read_excel("Metadata_for_COMPASS.xlsx", sheet = 2))

#take only one stim at a time, plus unstim
#Be careful here, this is hard-coded
metadata <- metadata[1:56,1:6]
colnames(metadata) <- metadata[1,]
metadata <- metadata[2:nrow(metadata),]
metadata$unique_id <- paste0(metadata[[4]], "_", metadata[[3]], "_", metadata[[1]])

#for ones with FMO or NA
ind <- c(which(grepl("FMO", metadata$`Mouse ID`)), 
         which(is.na(metadata$Mouse_ID)),
         which(grepl("uninf", metadata$Genotype)))
cd4_p300_data <- cd4_p300_data[-ind,]
cd4_unstim_data <- cd4_unstim_data[-ind,]
metadata <- metadata[-ind,]

#doing background correction on data (subtract unstim freq from stim freq)
corrected_dat <- cd4_p300_data - cd4_unstim_data
corrected_dat[corrected_dat<0] <- 0
rownames(corrected_dat) <- meta$unique_id
corrected_dat$Phenotype <- meta$Phenotype

#changing column names to a COMPASS format
nms = basename(colnames(corrected_dat))
nms = gsub(" \\| Freq. of Parent \\(\\%\\)", "", nms)
nms = gsub("\\.{3}\\d+", "", nms)
nms = COMPASS:::translate_marker_names(nms)
colnames(corrected_dat) = nms

#take only subsets previously determined to pass frequency filter in COMPASS
corrected_dat <- corrected_dat[,filtered_subsets]

#calculating p values and doing multiple hypothesis correction
ncan <- which(metadata$Phenotype == "R")
canon <- which(metadata$Phenotype == "C")

p <- c()
for(x in filtered_subsets){
  re <- corrected_dat[ncan, grep(paste0("^", x, "$"), colnames(corrected_dat))]
  ca <- corrected_dat[canon, grep(paste0("^", x, "$"), colnames(corrected_dat))]
  p <- c(p, wilcox.test(re, ca)[["p.value"]])
}
p <- p.adjust(p, method = "BH")
names(p) <- filtered_subsets

#plotting just the subsets that show significant differences between R and C
#(NCan and Canonical)
p_sig <- p[which(p < 0.05)]
p_sig <- names(p_sig)
d <- data.frame(metadata$Phenotype)
for(i in 1:length(p_sig)){
  d[p_sig[i]] <- corrected_dat[[grep(paste0("^", p_sig[i], "$"), colnames(corrected_dat))]]
}

d <- melt(d)

# Creating a data frame with p-values corresponding to each facet
p_value_data <- data.frame(
  variable = names(p[which(p < 0.05)]),
  p_value = p[which(p < 0.05)]
)

# Merging p-values into the main dataset to align with each facet
d_with_pvalues <- merge(d, p_value_data, by = "variable", all.x = FALSE, all.y = TRUE)

# Calculate midpoints for each variable
midpoints <- aggregate(value ~ variable + metadata.Phenotype, d_with_pvalues, median)

# Calculate midpoint for each variable between "R" and "C"
midpoints_summary <- aggregate(value ~ variable, midpoints, 
                               function(x) diff(range(x)) / 2 + min(x))

# Calculate midpoint between R and C for each variable
midpoints_R_C <- aggregate(value ~ variable, d_with_pvalues, 
                           function(x) diff(range(x)) / 2 + min(x))

# Merge midpoints with p-values for R and C
midpoints_p_values <- merge(midpoints_summary, midpoints_R_C, by = "variable")

#Now separate into 2 subsets for good visualization

# Filter variables with maximum value greater than 2
large_subset <- d %>%
  group_by(variable) %>%
  summarise(max_value = max(value)) %>%
  filter(max_value > 2)

# Filter variables with maximum value less than 2
small_subset <- d %>%
  group_by(variable) %>%
  summarise(max_value = max(value)) %>%
  filter(max_value < 2)

small_data <- d_with_pvalues[which(d_with_pvalues$variable %in% small_subset$variable),]
plot_small <- ggplot(data = small_data, 
                     aes(x = metadata.Phenotype, y = value)) +
  geom_jitter(aes(color = metadata.Phenotype, 
                  fill = metadata.Phenotype), size = 1) +
  stat_summary(fun = mean, geom = "crossbar", width = 0.4, linewidth = 0.2,
               aes(color = 'black')) +
  stat_summary(fun.data = mean_se, geom = "errorbar") +
  facet_grid(. ~ variable, scales = "free_x") +  # Facet on a single x-axis
  theme_classic() +
  scale_color_manual(breaks = c("R", "C"), values = c('#8C02CC', '#5192C3', 'black'))+
  scale_fill_manual(breaks = c("R", "C"), values = c('#8C02CC', '#5192C3', 'black'))+
  labs(x = "Phenotype", y = "Percentage responding CD4 T cells") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
         legend.position = "none") +
  # Adding p-values using annotate or geom_text
   geom_text(data = unique(subset(small_data, !is.na(p_value))), 
             aes(label = paste("p =", formatC(p_value, 
                                              digits = 3, format = "f"))),
            x = Inf, y = max(small_data$value) / 2, hjust = 1, vjust = 0.5, size = 3)

pdf("your_pdf.pdf", width = 5, height = 7)
print(plot_small)
dev.off()

large_data <- d_with_pvalues[which(d_with_pvalues$variable %in% large_subset$variable),]
plot_large <- ggplot(data = large_data, 
                     aes(x = metadata.Phenotype, y = value)) +
  geom_jitter(aes(color = metadata.Phenotype, 
                  fill = metadata.Phenotype), size = 1) +
  stat_summary(fun = mean, geom = "crossbar", width = 0.4, linewidth = 0.2,
               aes(color = 'black')) +
  stat_summary(fun.data = mean_se, geom = "errorbar") +
  facet_grid(. ~ variable, scales = "free_x") +  # Facet on a single x-axis
  theme_classic() +
  scale_color_manual(breaks = c("R", "C"), values = c('#8C02CC', '#5192C3', 'black'))+
  scale_fill_manual(breaks = c("R", "C"), values = c('#8C02CC', '#5192C3', 'black'))+
  labs(x = "Phenotype", y = "Percentage responding CD4 T cells") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
         legend.position = "none") +
  # Adding p-values using annotate or geom_text
   geom_text(data = unique(subset(large_data, !is.na(p_value))), 
             aes(label = paste("p =", formatC(p_value, 
                                              digits = 3, format = "f"))),
            x = Inf, y = max(large_data$value) / 2, hjust = 1, vjust = 0.5, size = 3)

pdf("your_pdf.pdf", width = 5, height = 7)
print(plot_large)
dev.off()
```

#Replace the list of cytokines below with the cytokines in your ICS data- that
#should be the only change you need to make
```{r table for cytokine combinations}

# Function to process multiple strings and create a combined table
process_strings_combined <- function(strings) {
  cytokines <- c("CD40L", "CD107a", "IFNg", "IL2", "IL4", "IL10", "IL17", "TNF")
  combined_table <- data.frame(Cytokine = cytokines)
  
  for (i in seq_along(strings)) {
    cytokine_values <- unlist(strsplit(strings[i], "&"))
    #don't take any blank spaces or cytokines not in input list
    combined_string <- paste(cytokines, collapse = "|")
    cytokine_values <- cytokine_values[which(grepl(combined_string, cytokine_values))]
    print(cytokine_values)
    values <- rep("+", length(cytokines))  # Initialize values for each cytokine
    
    for (j in 1:length(cytokine_values)) {
      if (grepl("^!|^&!", cytokine_values[j])) {
        values[j] <- " -"  # Update values based on patterns
      }
    }
    
    col_name <- paste(i)
    combined_table[[col_name]] <- values
  }
  
  return(combined_table)
}

# Process strings and create a combined table
result_combined_table <- process_strings_combined(p_sig)

#Small data

# Process strings and create a combined table
small_names <- as.character(unique(small_subset$variable))
result_combined_table <- process_strings_combined(small_names)
# Create the table grob without background lines and shades
table_grob <- tableGrob(result_combined_table, 
                        theme = ttheme_minimal(
                          base_size = 8,
                          padding = unit(c(5, 5), "mm"), # Adjust padding here
                          core = list(bg_params = list(fill = NA)),
                          colhead = list(bg_params = list(fill = NA)),
                          rowhead = list(bg_params = list(fill = NA))
                        ))
# Arrange plot and table
combined <- grid.arrange(plot_small, table_grob, ncol = 1,
                         heights = c(2, 1))

                         #heights = unit(c(1, 1), "null"))
# Set up the file name
pdf_file <- "COMPASS_sig_percentages_withTable_small_se.pdf"
# Save the combined plot and table to a PDF file
ggsave(pdf_file, combined, width = 8, height = 7)

#Large data

# Process strings and create a combined table
large_names <- as.character(unique(large_subset$variable))
result_combined_table <- process_strings_combined(large_names)# Create the table grob without background lines and shades
table_grob <- tableGrob(result_combined_table, 
                        theme = ttheme_minimal(
                          base_size = 8,
                          padding = unit(c(5, 5), "mm"), # Adjust padding here
                          core = list(bg_params = list(fill = NA)),
                          colhead = list(bg_params = list(fill = NA)),
                          rowhead = list(bg_params = list(fill = NA))
                        ))
# Arrange plot and table
combined <- grid.arrange(plot_large, table_grob, ncol = 1,
                         heights = c(2, 1))

                         #heights = unit(c(1, 1), "null"))
# Set up the file name
pdf_file <- "COMPASS_sig_percentages_withTable_large_se.pdf"
# Save the combined plot and table to a PDF file
ggsave(pdf_file, combined, width = 2.85, height = 7) 
#Change the width proportionally to # cytokine subsets plotted, to keep width same
#across different stimulations

```

